---
title: "Reproducible Research - Peer Assessment 1"
output: html_document
---


  ----

### Load Required Libraries  
```{r Libraries, echo=TRUE}
    library(lattice)
    library(plyr)
```  
  
  ----
  
### Loading and preprocessing the data  

* Read data into data frame  
  
```{r Read Dataset, echo=TRUE, }
    data <- read.csv(paste(c(getwd(),"/activity.csv"),collapse=""))
```


  * Convert date variable to date format  
  
```{r DT-Convert date, echo=TRUE}
    data$date <- as.Date(data$date)
```


* Create subset with observations containing NA values removed  
  
```{r DT-Remove NAs, echo=TRUE}
    amdata <- data[complete.cases(data),]
```

  ----

### Mean total number of steps taken per day  


  * Make a histogram of the total number of steps taken each day     
  
```{r Mean Steps-Histogram, echo=TRUE}
    stepsbyday<- ddply(amdata, c("date"), function (x) apply(x[1], 2, sum))
    hist(stepsbyday$steps,main="Histogram of Total Steps per Day",xlab=
           "Total Number of Steps per Day",ylab="Frequency",col='purple')    
```


  * Report the mean total number of steps taken each day  
  
```{r Mean Steps-Mean, echo=TRUE}
    origmean <- as.integer(mean(stepsbyday$steps))
```
    The mean of the dataset is `r origmean`


  * Report the median total number of steps taken each day  
  
```{r Mean Steps-Median, echo=TRUE}
    origmed <- median(stepsbyday$steps)
```
    The median of the dataset is `r origmed`  

  ----

### Average daily activity pattern  


  * Time series plot of the 5-minute interval and average number of steps taken   
    averaged across all days  

```{r Time-Series Plot, echo=TRUE}
    stepsbyinterval <- aggregate(steps ~ interval, data=amdata, FUN=mean)
    plot(stepsbyinterval, type="l",xlab="5-Minute Interval", ylab=
           "Average Steps",col="red")
```  


  * Which 5-minute interval, on average across all the days in the dataset, 
    contains the maximum number of steps?   
    
```{r Max average steps, echo=TRUE}
      intwithmax <- stepsbyinterval$interval[which.max(stepsbyinterval$steps)]
```
    The interval `r intwithmax` contains the maximum number of steps  

  ----

### Imputing Missing Values   


  * Calculate and report the total number of missing values in the dataset 
    (ie, the total number of rows with NAs)  
    
```{r Missing Values, echo=TRUE}
      numNAs <- sum(is.na(data))
``` 
    The total number of observations with NA values is: `r numNAs`  
    
    
  * Devise a strategy for fill in all of the missing values in the dataset
     - Strategy is to replace NAs with average for the corresponding interval    
      
     --- 
  * Create a new dataset that is equal to the original dataset but with the 
    missing data filled in    
      
      
```{r Fill missing values, echo=TRUE}
    ##Get averages for full dataset (with NA observations)
    fullstepsbyinterval <- aggregate(steps ~ interval, data=amdata, FUN=mean)
    
    ##Merge the orginal dataset with the average interval values 
    newdataset <- merge(data, fullstepsbyinterval, by="interval")

    ##Update NA steps.x to steps.y 
    for (i in 1:nrow(newdataset))
      {
      if (is.na(newdataset[i,]$steps.x))
        {
          newdataset[i,]$steps.x <- round(newdataset[i,]$steps.y,digits=0)
        }
      }
  
    ##Remove columns not required
    newdataset <- newdataset[,1:3]

    ##Reset columns names to original 
    colnames(newdataset) <- c("interval","steps","date")

    ##Change column types to original 
     newdataset$steps <- as.integer(newdataset$steps)
```

The new dateset contains `r nrow(newdataset)` rows:  
    - `r sum(is.na(newdataset))` observations contain NA values 


* Make a histogram of the total number of steps each day with the imputed data  
  
```{r Mean Steps-Histogram (imputed data), echo=TRUE}
    stepsbyday2<- ddply(newdataset, c("date"), function (x) apply(x[2], 2, sum))
    hist(stepsbyday2$steps,main="Histogram of Total Steps per Day (new data)",
         xlab="Total Number of Steps per Day",ylab="Frequency",col='purple')    
```  


  * Report the mean total number of steps taken each day with the imputed data  
  
```{r Mean Steps-Mean(imputed data), echo=TRUE}
    newmean <- as.integer(mean(stepsbyday2$steps))
```  
  The mean of the imputed dataset is `r newmean`  
  
  * Report the median total number of steps taken each day with the imputed data    
    
```{r Mean Steps-Median(imputed data), echo=TRUE}
     newmed <- median(stepsbyday2$steps)
```  
  The median of the imputed dataset is `r newmed`  
  
  
  * Do these values differ from the original estimates? What is the impact of 
    imputing missing data on the estimate of the total daily number of steps?   
    
    - The mean of the two datasets is the same 
    - The median is slightly lower than the original 
    - The histograms have a relatively consistent shape, however, the frequency 
      is higher in the imputed data.    

  ----

### Activity patterns on weekdays and weekends


  * Create a new factor variable in the dateset with two levels - "weekday" and 
    "weekend"  
    
```{r Add weekday/weekend variable, echo=TRUE}
    ##Add empty variable for type of day
    newdataset$typeday <- ""
    
    ##Add type of day   
    for (i in 1:nrow(newdataset))
    {
      if(weekdays(newdataset[i,]$date) %in% c("Saturday", "Sunday"))
      {
        newdataset[i,]$typeday <- "Weekend"
      }
        
      else newdataset[i,]$typeday <- "Weekday"
    
    }
    
    ##Convert to factor
    newdataset$typeday <- as.factor(newdataset$typeday)
```


  * Make a panel plot containing a time series plot (ie, type = "l") of the 
    5-minute interval and the average number of steps taken, averaged across all
    weekdays/weekend days  
    
```{r Panel Plots, echo=TRUE }    
  
  ##Create dataset with average by typeday and interval
  stepsbydaytype<- ddply(newdataset, c("interval","typeday"), 
                         function (x) apply(x[2], 2, sum))
  ##Plot in 2 panels 
  xyplot(steps~interval|factor(typeday),data=stepsbydaytype,aspect=1/2,type="l")
```

  ----
  